scRNAseq

### 2018-02-26 datasets/scrnaseq/scRNAseq-keegan.Rmd

For data processing, we follow practices and recommendations from Soneson et al., 2017. Briefly, we limit ourselves to 2-group comparisons, and  select the maximal size defined as the smallest number of cells in either group as the group-level sample size for the DE analyses. We do not filter genes as filtering may affect the subsequent false discovery control steps.

@@

# subset by groups of interest
humandata <- humandata[, (colData(humandata)$source_name_ch1 %in% 
                         c("Somatic Cells", "Primordial Germ Cells")) &
                         (colData(humandata)$characteristics_ch1 == 
                         c("developmental stage: 7 week gestation"))]
cdata=colData(humandata)

@@

# look at number of cells in each group
table(colData(humandata)$source_name_ch1)

@@

# subset by groups of interest
humandata2 <- humandata2[, (colData(humandata2)$characteristics_ch1.7 %in% 
                         c("neoplastic: Neoplastic", "neoplastic: Regular")) &
                         (colData(humandata2)$characteristics_ch1.5 %in% 
                         c("tsne cluster: 11",  "tsne cluster: 8"))]
cdata=colData(humandata2)

@@

# look at number of cells in each group
table(colData(humandata2)$characteristics_ch1.7)

@@

# subset by groups of interest
mousedata <- mousedata[, colData(mousedata)$source_name_ch1 %in% 
                         c("16-cell stage blastomere", 
                           "Mid blastocyst cell (92-94h post-fertilization)")]
cdata=colData(mousedata)

@@

# look at number of cells in each group
table(as.character(colData(mousedata)$source_name_ch1))

@@

# subset by groups of interest
mousedata2 <- mousedata2[, (colData(mousedata2)$characteristics_ch1.2 ==
                              "condition: No stimulation") |
                           (colData(mousedata2)$characteristics_ch1.2 ==
                              "condition: LPS" &
                            colData(mousedata2)$characteristics_ch1.1 %in% 
                              "time point: 150 min") ]
cdata=colData(mousedata2)

@@

# look at number of cells in each group
table(as.character(colData(mousedata2)$characteristics_ch1.2))

@@

** Still need to clean up from here on; nothing below is eval'd**

@@

We will first assign names for output files of the DE methods as they will be handy downstream.

```{r, filenames, eval=FALSE}
h_wilcox_dat1=paste0(resdir,"h_data_wilcox_p_GSE63818-GPL16791.rds")
h_wilcox_dat2=paste0(resdir,"h_data_wilcox_p_GSE84465.rds")
m_wilcox_dat1=paste0(resdir,"m_data_wilcox_p_GSE45719.rds")
m_wilcox_dat2=paste0(resdir,"m_data_wilcox_p_GSE94383.rds")

h_scdd_dat1=paste0(resdir,"humanplate_scDatEx_ks_GSE63818-GPL16791.rds")
h_scdd_dat2=paste0(resdir,"humanplate_scDatEx_ks_GSE84465.rds")
m_scdd_dat1=paste0(resdir,"mouseplate_scDatEx_ks_GSE45719.rds")
m_scdd_dat2=paste0(resdir,"mouseplate_scDatEx_ks_GSE94383.rds")

m_mast_dat1=paste0(resdir,"mouseplate_masttpm_GSE45719.rds")
h_mast_dat1=paste0(resdir,"humanplate_masttpm_GSE63818-GPL16791.rds")
h_mast_dat2=paste0(resdir,"humanplate_masttpm_GSE84465.rds")
m_mast_dat2=paste0(resdir,"mouseplate_masttpm_GSE94383.rds")

@@

```{r, wilcoxon, eval=FALSE}

g_data_sub=readRDS(paste0(datdir,"h_g_nofilter_subdata_GSE63818-GPL16791.rds"))
groups=c(rep(1,(dim(g_data_sub)[2]/2)),rep(2,(dim(g_data_sub)[2]/2)))
compute_wilcoxde_sc(g_data_sub,h_wilcox_dat1,groups)

g_data_sub=readRDS(paste0(datdir,"h_g_nofilter_subdata_GSE84465.rds"))
groups=c(rep(1,(dim(g_data_sub)[2]/2)),rep(2,(dim(g_data_sub)[2]/2)))
compute_wilcoxde_sc(g_data_sub,h_wilcox_dat2,groups)

g_data_sub=readRDS(paste0(datdir,"m_g_nofilter_subdata_GSE45719.rds"))
groups=c(rep(1,(dim(g_data_sub)[2]/2)),rep(2,(dim(g_data_sub)[2]/2)))
compute_wilcoxde_sc(g_data_sub,m_wilcox_dat1,groups)

g_data_sub=readRDS(paste0(datdir,"m_g_nofilter_subdata_GSE94383.rds"))
groups=c(rep(1,(dim(g_data_sub)[2]/2)),rep(2,(dim(g_data_sub)[2]/2)))
compute_wilcoxde_sc(g_data_sub,m_wilcox_dat2,groups)

### Differences in distributions of gene expression in the 2 conditions

We apply scDD to log-transformed counts as below. Th helper functions set up the Single-Cell experiment object.

>>>>

For data processing, we follow practices and recommendations from Soneson et al., 2017. Briefly, we limit ourselves to 2-group comparisons. We do not filter genes as filtering may affect the subsequent false discovery control steps.

@@

# subset by groups of interest
humandata <- humandata[, (colData(humandata)$source_name_ch1 %in% 
                         c("Somatic Cells", "Primordial Germ Cells")) &
                         (colData(humandata)$characteristics_ch1 == 
                         c("developmental stage: 7 week gestation"))]
cdata=colData(humandata)
cdata$group <- cdata$source_name_ch1

@@

# look at number of cells in each group
table(colData(humandata)$group)

@@

# subset by groups of interest
humandata2 <- humandata2[, (colData(humandata2)$characteristics_ch1.7 %in% 
                         c("neoplastic: Neoplastic", "neoplastic: Regular")) &
                         (colData(humandata2)$characteristics_ch1.5 %in% 
                         c("tsne cluster: 11",  "tsne cluster: 8"))]
cdata=colData(humandata2)
cdata$group <- cdata$characteristics_ch1.7

@@

# look at number of cells in each group
table(colData(humandata2)$group)

@@

# subset by groups of interest
mousedata <- mousedata[, colData(mousedata)$source_name_ch1 %in% 
                         c("16-cell stage blastomere", 
                           "Mid blastocyst cell (92-94h post-fertilization)")]
cdata=colData(mousedata)
cdata$group <- cdata$source_name_ch1

@@

# look at number of cells in each group
table(as.character(colData(mousedata)$group))

@@

# subset by groups of interest
mousedata2 <- mousedata2[, (colData(mousedata2)$characteristics_ch1.2 ==
                              "condition: No stimulation") |
                           (colData(mousedata2)$characteristics_ch1.2 ==
                              "condition: LPS" &
                            colData(mousedata2)$characteristics_ch1.1 %in% 
                              "time point: 150 min") ]
cdata=colData(mousedata2)
cdata$group <- cdata$characteristics_ch1.2

@@

# look at number of cells in each group
table(as.character(colData(mousedata2)$group))

@@

@@

@@

```{r, wilcoxon}
# function to add wilcox p-values
compute_wilcoxde_sc <- function(dat){
  expr <- assays(dat)[["TPM"]] 
  tmm <- edgeR::calcNormFactors(expr)
  expr <- edgeR::cpm(expr, lib.size = tmm * colSums(expr))
  rowData(dat)$wilcox_p <- sapply(seq_len(nrow(expr)), 
    function(i) {
      wilcox.test(expr[i, ] ~ colData(dat)$group)$p.value
    })
  return(dat)
}

humandata <- compute_wilcoxde_sc(humandata)

humandata2 <- compute_wilcoxde_sc(humandata2)

mousedata <- compute_wilcoxde_sc(mousedata)

mousedata2 <- compute_wilcoxde_sc(mousedata2)
```

### Differences in distributions of gene expression in the 2 conditions
** Still need to clean up from here on; nothing below is eval'd**
We apply scDD to log-transformed counts as below. Th helper functions set up the Single-Cell experiment object.

### 2018-02-26 datasets/scrnaseq/scRNAseq-latest.Rmd

#installation
# source("https://bioconductor.org/biocLite.R")
# biocLite("swfdr")
# biocLite("IHW")
# biocLite("qvalue")
# install.packages("ashr")
# install.packages("fdrtool")
# biocLite("edgeR")
# source("https://bioconductor.org/biocLite.R")
# biocLite("scran")
#  Samseq installation
# source("https://bioconductor.org/biocLite.R")
# biocLite("impute") 
# install.packages("samr")
# source("https://bioconductor.org/biocLite.R")
#biocLite("MultiAssayExperiment")
#source("https://bioconductor.org/biocLite.R")
#biocLite("SummarizedExperiment")
#biocLite("BiocGenerics")
#biocLite("S4Vectors")
#biocLite("BiocParallel")
#install_github("areyesq89/SummarizedBenchmark")
#biocLite("EBSeq")
#biocLite("scDD")
#source("https://bioconductor.org/biocLite.R")
#biocLite("SingleCellExperiment")
#source("https://bioconductor.org/biocLite.R")
#biocLite("MAST")
#install.packages("FDRreg")

# packages for data analyses
suppressMessages(library(MultiAssayExperiment))
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(cowplot))
suppressMessages(library(UpSetR))
suppressMessages(library(hexbin))

# packages for differential expression testing
suppressMessages(library(impute))
#suppressMessages(library(samr)
suppressMessages(library(scDD))
suppressMessages(library(EBSeq))
suppressMessages(library(edgeR))
suppressMessages(library(MAST))

# benchmark methods
suppressMessages(library(IHW))
suppressMessages(library(ashr))
suppressMessages(library(qvalue))
suppressMessages(library(swfdr))
suppressMessages(library(fdrtool))
suppressMessages(library(FDRreg))

# summarizebenchmark
suppressMessages(library(SummarizedBenchmark))
suppressMessages(library(SingleCellExperiment))

# useful libraries
suppressMessages(library(R.utils))
suppressMessages(library(knitr))
suppressMessages(library(BiocParallel))
core <- as.numeric(Sys.getenv("SLURM_NTASKS"))
multicoreParam <- MulticoreParam(workers = cores)

@@

```{bash, humandata-download}

download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE63818-GPL16791.rds",destfile=paste0(datdir,"GSE63818-GPL16791.rds"))

download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE84465.rds",destfile=paste0(datdir,"GSE84465.rds"))

@@

```{bash, mousedata-download}

download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE45719.rds",destfile=paste0(datdir,"GSE45719.rds"))

download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE94383.rds",destfile=paste0(datdir,"GSE94383.rds"))

@@

```{r, nofilter1}
g_data=cbind(g1_data,g2_data)
g_data_sub=g_data[rowSums(g_data==0)!=dim(g_data)[2],] 
saveRDS(g_data_sub,paste0(datdir,"h_g_nofilter_subdata_GSE63818-GPL16791.rds")) # removing rows with all zeros reduces the number of genes to 33169

@@

saveRDS(g1_ids,paste0(datdir,"g1_ids_GSE63818-GPL16791.rds"))
saveRDS(g2_ids,paste0(datdir,"g2_ids_GSE63818-GPL16791.rds"))
```

@@

```{r, nofilter2}
g_data=cbind(g1_data,g2_data)
g_data_sub=g_data[rowSums(g_data==0)!=dim(g_data)[2],] 
saveRDS(g_data_sub,paste0(datdir,"h_g_nofilter_subdata_GSE84465.rds")) # removing rows with all zeros reduces the number of genes to 52942

g_cdata=cbind(g1_cdata,g2_cdata)
g_cdata_sub=g_cdata[rowSums(g_cdata==0)!=dim(g_cdata)[2],]
saveRDS(g_cdata_sub,paste0(datdir,"h_g_nofilter_subcdata_GSE84465.rds")) # removing rows with all zeros reduces the number of genes to 52942

saveRDS(g1_ids,paste0(datdir,"g1_ids_GSE84465.rds"))
saveRDS(g2_ids,paste0(datdir,"g2_ids_GSE84465.rds"))
```

@@

```{r,mouse}
##mouse data
mousedata=readRDS(paste0(datdir,"GSE45719.rds"))
#get gene-level data
genes=experiments(mousedata)[1]
#TPMS
#get tpms
tpms=assays(genes$gene)[1]
#get tpm dataframe
mouse_tpms=tpms$TPM
# get column data/metadata for sample
cdata=colData(mousedata)
cdata=as.data.frame(cdata)
# select groups of interest
g1 = cdata%>%subset(source_name_ch1=="16-cell stage blastomere")
g2 = cdata%>%subset(source_name_ch1=="Mid blastocyst cell (92-94h post-fertilization)")
# select maximal sample size
sample_size = c(dim(g1)[1],dim(g2)[1])
ssize = which.min(sample_size)
g1_ids=sample.int(dim(g1)[1], size = sample_size[ssize], replace = FALSE)
g2_ids=sample.int(dim(g2)[1], size = sample_size[ssize], replace = FALSE)
# get group data of maximal size
g1_data=mouse_tpms[,colnames(mouse_tpms)%in%g1$geo_accession[g1_ids]]
g2_data=mouse_tpms[,colnames(mouse_tpms)%in%g2$geo_accession[g2_ids]]
#save data
saveRDS(g1_data,paste0(datdir,"m_g1_data_GSE45719.rds"))
saveRDS(g2_data,paste0(datdir,"m_g2_data_GSE45719.rds"))
#COUNTS
#get counts
count=assays(genes$gene)[2]
#get count dataframe
mouse_count=count$count
# get column data/metadata for sample
cdata=colData(mousedata)
cdata=as.data.frame(cdata)
# get group data of maximal size
g1_cdata=mouse_count[,colnames(mouse_count)%in%g1$geo_accession[g1_ids]]
g2_cdata=mouse_count[,colnames(mouse_count)%in%g2$geo_accession[g2_ids]]
#save data
saveRDS(g1_cdata,paste0(datdir,"m_g1_cdata_GSE45719.rds"))
saveRDS(g2_cdata,paste0(datdir,"m_g2_cdata_GSE45719.rds"))

>>>>

# packages for data analyses
library(MultiAssayExperiment)
library(dplyr)
library(ggplot2)
library(cowplot)
library(UpSetR)
library(hexbin)

# packages for differential expression testing
library(impute)
library(scDD)
library(EBSeq)
library(edgeR)
library(MAST)

# summarizebenchmark
library(SummarizedBenchmark)
library(SingleCellExperiment)

# useful libraries
library(R.utils)
library(knitr)
library(BiocParallel)

cores <- as.numeric(Sys.getenv("SLURM_NTASKS"))
multicoreParam <- MulticoreParam(workers = cores)

@@

```{r, humandata-download}
# all data and results are on the wd: /net/irizarryfs01/srv/export/irizarryfs01_backed_up/share_root/fdrbenchmark-scrnaseq

#these look well separated on the tsne plots on the conquer manuscript supplement
download.file("http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE63818-GPL16791.rds",
  destfile=paste0(datdir, "GSE63818-GPL16791.rds"))

## human ds2
download.file("http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE84465.rds",
  destfile=paste0(datdir, "GSE84465.rds"))

@@

```{r, mousedata-download}
#another well-separated dataset on tsne: GSE60749−GPL13112
download.file("http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE45719.rds",
  destfile=paste0(datdir, "GSE45719.rds"))

download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE94383.rds",
  destfile=paste0(datdir,"GSE94383.rds"))
```

@@

```{r, nofilter1}
g_data=cbind(g1_data,g2_data)
g_data_sub=g_data[rowSums(g_data==0)!=dim(g_data)[2],] 
saveRDS(g_data_sub, paste0(datdir, "h_g_nofilter_subdata_GSE63818-GPL16791.rds")) # removing rows with all zeros reduces the number of genes to 33169

@@

saveRDS(g1_ids, paste0(datdir, "g1_ids_GSE63818-GPL16791.rds"))
saveRDS(g2_ids, paste0(datdir, "g2_ids_GSE63818-GPL16791.rds"))
```

@@

```{r, nofilter2}
g_data=cbind(g1_data,g2_data)
g_data_sub=g_data[rowSums(g_data==0)!=dim(g_data)[2],]

saveRDS(g_data_sub, paste0(datdir, "h_g_nofilter_subdata_GSE84465.rds")) # removing rows with all zeros reduces the number of genes to 52942

g_cdata=cbind(g1_cdata,g2_cdata)
g_cdata_sub=g_cdata[rowSums(g_cdata==0)!=dim(g_cdata)[2],]
saveRDS(g_cdata_sub, paste0(datdir, "h_g_nofilter_subcdata_GSE84465.rds")) # removing rows with all zeros reduces the number of genes to 52942

saveRDS(g1_ids, paste0(datdir, "g1_ids_GSE84465.rds"))
saveRDS(g2_ids, paste0(datdir, "g2_ids_GSE84465.rds"))
```

@@

```{r,mouse}
##mouse data
mousedata=readRDS(paste0(datdir, "GSE45719.rds"))

#get gene-level data
genes=experiments(mousedata)[1]
#TPMS
#get tpms
tpms=assays(genes$gene)[1]
#get tpm dataframe
mouse_tpms=tpms$TPM
# get column data/metadata for sample
cdata=colData(mousedata)
cdata=as.data.frame(cdata)
# select groups of interest
g1 = cdata%>%subset(source_name_ch1=="16-cell stage blastomere")
g2 = cdata%>%subset(source_name_ch1=="Mid blastocyst cell (92-94h post-fertilization)")
# select maximal sample size
sample_size = c(dim(g1)[1],dim(g2)[1])
ssize = which.min(sample_size)
g1_ids=sample.int(dim(g1)[1], size = sample_size[ssize], replace = FALSE)
g2_ids=sample.int(dim(g2)[1], size = sample_size[ssize], replace = FALSE)
# get group data of maximal size
g1_data=mouse_tpms[,colnames(mouse_tpms)%in%g1$geo_accession[g1_ids]]
g2_data=mouse_tpms[,colnames(mouse_tpms)%in%g2$geo_accession[g2_ids]]
#save data
saveRDS(g1_data,paste0(datdir, "m_g1_data_GSE45719.rds"))
saveRDS(g2_data,paste0(datdir, "m_g2_data_GSE45719.rds"))
#COUNTS
#get counts
count=assays(genes$gene)[2]
#get count dataframe
mouse_count=count$count
# get column data/metadata for sample
cdata=colData(mousedata)
cdata=as.data.frame(cdata)
# get group data of maximal size
g1_cdata=mouse_count[,colnames(mouse_count)%in%g1$geo_accession[g1_ids]]
g2_cdata=mouse_count[,colnames(mouse_count)%in%g2$geo_accession[g2_ids]]
#save data
saveRDS(g1_cdata,paste0(datdir,"m_g1_cdata_GSE45719.rds"))
saveRDS(g2_cdata,paste0(datdir,"m_g2_cdata_GSE45719.rds"))
